Measuring contrast processing in the visual system using SSVEP

### 2024-11-27 responseToReviews.md

Two-limbed contrast response functions. Campbell and Maffei (J. Physiol,1970) in the earliest measurements of the VEP contrast response function reported that it sometimes had two “limbs” – one at low contrast and a second one with a steeper slope at higher contrasts.  Two-limbed contrast response functions have also been found by Nakayama and Mackeben (Vision Res,1982) in macaque with electrodes on the dura, by Apkarian and Tyler (Vision Res, 1985),  Norcia et al., (Vision Res,1989) and Souza et al., (IOVS, 2007), to name a few.  As noted, the sigmoidal model assumes that the population response is unitary, but the two-limbed functions suggest that it is a mixture of populations. The two limbs may not be apparent when the function is sampled with only a few points over a wide range of contrast.

Harmonics with different CRFs.  Bobak et al., (IOVS, 1986), found that the first harmonic (1f) and second harmonic (2f) of the pattern onset/offset SSVEP had different thresholds but similar slopes once above threshold. Norcia et al., (TVST, 2024) found different thresholds and slope for 1f and 2f responses driven by contrast modulated dynamic noise. This again suggests the presence of multiple underlying populations.  The behavior also raises the question of whether to pool harmonics when estimating the response function. This was alluded to in the simulation, but could be expanded on.

The lack of dynamics in the normalization model, as presented should be discussed, the brief mention of Tsai et al. aside. The authors may wish to cite another framework that explicitly models dynamics (e.g. amplitude and phase) in the context of the contrast response function (Zemon and Gordon, Vision Res, 2006) .

### Response
> We thank the reviewer for these positive comments. In general this was intended to be an overview in the sense of a 'primer' for SSVEP rather than a detailed historical account of the field - hence the inclusion of code examples and relatively domain-agnostic language. However, the reviewer raises some important general issues and we now address them: Specifically, we include references to the idea that SSVEP can engage more than one type of neuron, that the population responses is not trivially derived from what is known about the properties of individual neurons and that fitting with a sigmoidal response function may mask some complexities in the CRF that are more readily seen with denser sampling. We think the Bobak reference is IOVS 1984 and we have also included the Kaestner TVST 2024 reference (thank you for drawing our attention to both of these).

@@

### Specific comments

Adaptation effects.  Adapted contrast response functions have been measured by Nelson et al., (EEG Journal, 1984) by comparing CRFs measured by sweeping contrast from low to high vs high to low. They found that responses at low contrast were smaller in the “down” sweeps and attributed it to differences in prior adaptation.

> we could not find this reference. We did find an excellent (and sometimes humorous) review of SSVEPs by Nelson (Nelson, J. I., Seiple, W. H., Kupersmith, M. J., & Carr, R. E. (1984). Lock-in Techniques for the Swept Stimulus Evoked Potential. Journal of Clinical Neurophysiology, 1(4), 409–436. doi:10.1097/00004691-198401040-00002) but could not find a mention of adaptation effects.

Other studies also measured adaptation effects on the CRF Heinrich and Bach, (IOVS, 2001), Ross and Speed (1991, Proc Roy Soc). And the hard-to-find Bach, Greenlee and Buhler (Clin Vis Sci, 1996) paper that modeled the effect within a contrast gain control framework.

> The Ross and Speed paper measures masking using SSVEP but other adaptation measures are, as far as we can tell, psychophysical.  Heinrich and Bach are concerned with transient changes in SSVEP amplitude within a recording which may have an adaptation origin but which manifest as response increases over the first few seconds and seems a little mysterious. Bach, Greenlee and Buhler is, however, pretty unequivocal and we now include it. We like the suggestion that the  'increase' in response is attributableto a shift from a square wave response to a more sinusoidal waveform as the apparent contrast of the stimulus is decreased by adaptation-driven gain control. The response power which was distributed across odd harmonics in the original response can now be concentrated in the 1st harmonic after adaptation.

Gain control development.  Morrone and Burr (Nature, 1986) used frequency tagging to measure orientation dependence of masking in infants and adults, as did Candy et al. The difference in the two studies is that in adults Morrone and Burr found response gain for cross-oriented maskers but Candy et al found contrast gain.   Skoczenski and Norcia used a high contrast 30 Hz dynamic 2D noise masker and variable contrast reversing gratings. They found masking consistent with a contrast gain effect.

### 2024-11-27 review.qmd
Editing both paper and R2R - now at end of R1/ adaptation

---
title: "Measuring contrast processing in the visual system using SSVEP"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format:
  docx: 
    omitFigures: true
  pdf:

    keep-tex: true

@@

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. SSVEP signals can also be elicited by periodic changes of stimulus properties other than achromatic and chromatic contrast, such as motion, stereo depth, and facial identity or expression [see @Norcia2015, for an overview]; however our focus here is on the contrast response.

@@

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. The SSVEP response to a contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. For example, @Tsai2012 demonstrated that a gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. Our more recent work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.

>>>>

---
title: "Measuring contrast processing in the visual system using SSVEP"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format:
  #docx: 
    #omitFigures: true
  pdf: default
    #keep-tex: true

@@

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. Because they can arise from different neuronal computations, different populations and even different stages of the visual system, different harmonics can exhibit different input-output functions. For example, 1F and 2F responses can have different thresholds [@BobakIOVS1984] and scalp topographies [@reganEvokedPotentialsSpecific1973] while Kaester et al [@KaestnerTVST2024] show both different CRF slopes and thresholds for 1F and 2F components generated by dynamic noise.

SSVEP signals can also be elicited by periodic changes of stimulus properties other than achromatic and chromatic contrast, such as motion, stereo depth, and facial identity or expression [see @Norcia2015, for an overview]; however our focus here is on the contrast response.

@@

Although the SSVEP does seem to reflect the activity of relatively well-tuned neuronal populations , changing some aspect of the stimulus may change the nature of those populations. For example, Campbell and Maffei [@Campbell1970] noted that densely-sampled measurements of contrast responses can reveal the presence of two, qualitatively different types of neurons that exhibit different log-linear contrast functions (see also [@Tyler1985, @Nakayama1982, @Norcia1989 and @Souza2007]). Modeling such two-limbed contrast response functions using a single sigmoidal function is therefore an approximation.

@@

Even without considering a spatial component, the early visual system is far more complex than the model here suggests. For example as well as cells that code positive or negative contrast in a more or less continuous manner, the retina also contains 'transient' cells that code temporal changes in contrast. These cells [@Kuffler1957; @Alpern1971], and cells with similar properties in the LGN [@Levitt2001] and cortex [@Movshon1975; @Hubel1959], will introduce second harmonic components even when the stimulus itself is modulated in an on-off fashion. Analogously, in the spatial domain, so-called 'simple cells' are sensitive to the polarity of a spatial contrast modulation while 'complex cells' respond to the presence of patterned spatial contrast irrespective of its spatial position [@Hubel1962]. Both the amplitude and the phase of the SSVEP response to a single contrast-reversing sine-wave grating therefore contains information about nonlinear computations performed across a range of retinal and cortical cell types.

The complexity of even a simple simulation of the frequency-domain signal arising from non-linear interactions is intriguing. Presumably, the signal measured from early visual cortex is the result of a cascade of nonlinear retinal and cortical operations up to that point. It therefore contains a 'signature' or 'fingerprint' of the computational nature, order and parameters of those operations - including the shape of the transducer functions and the computations involved in signal combination. In principle, that information could be recovered from the SSVEP signal - a possibility recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is challenging, careful parametric variation of the input stimuli does allow us to fit models of early visual visual processing (e.g. [@Zemon2006Dynamic]) and, by incorporating mathematical models of gain control mechanisms with either instantaneous, or temporally-integrated gain control, it is possible to model gain control using SSVEP data. For example, [@Tsai2012] demonstrated that a relatively simple gain control model gave a good account of the pattern of intermodulation responses produced by two overlaid patterns flickering at different frequencies. This was achieved by passing full stimulus waveforms through the transducer nonlinearity, and calculating the Fourier spectrum of the model output. Our more recent work on signal combination across eyes and space similarly demonstrated close correspondence between the predictions of a computational model and empirical data in humans [@Baker2017]. More detailed modelling of intracortical recordings [@Groen2022] has revealed details of the timecourse of gain control effects, specifically that normalization is delayed slightly relative to the initial visual response.
